---
title: "Linking Ethnic Data from Africa"
output:
   md_document

---

The LEDA package contains a full pipeline to link
ethnic datasets from Africa. The main strength of LEDA
consists in leveraging the structure of the language
tree to provide a flexible link between any two ethnic
group that are linked to the tree.

The package allows lists of ethnic groups to be linked to each
other using three main linkage types: binary linking based
on the relations of sets of language nodes associated with
two groups; binary linking based on lingustic distances;
and a full computation of dyadic linguistic distances.

Usage of a LEDA object is structured around
*lists of ethnic groups*. These lists of groups
stem from the original datasets that have been joined
to the language tree. Lists are structured by data source,
country, year, or, in the case of survey data, survey rounds.
Via the language tree, any two lists of ethnic groups
can be linked to each other.


For full information on the LEDA project and methodology,
read the [paper](https://github.com/carl-mc/LEDA/raw/master/docs/LEDA_paper.pdf).

When using the LEDA package, please cite:
Müller-Crepon, Carl, Yannick Pengl, and Nils-Christian Bormann (2020).
*Linking Ethnic Data from Africa*. Unpublished working paper.


## Installation

You can directly download and install the LEDA package from GitHub.

```{r, eval = F}
library(devtools)
install_github(repo = "carl-mc/LEDA")
```


## Initialize linking object

The LEDA package is programmed in an object oriented manner. Once you initialize a LEDA-object, methods are applied directly to the object and either change the object or return the results of a query. See the documentation of the R-package [R6](https://cran.r-project.org/web/packages/R6/R6.pdf) for details.

### Create LEDA objects

```{r}
library(LEDA)
leda <- LEDA$new()
```


### Help files

Because all functionalities of the LEDA package are methods of LEDA objects, all documentation can be accessed by calling `?LEDA`.



### Datasets included in LEDA

To get a first overview of the possibilities coming with LEDA, start querying the 'list dictionary', which contains all metadata of all lists of ethnic groups that the LEDA project links to the Ethnologue language tree. Lists are identified by their country, the type of dataset (e.g. EPR, Afrobarometer, DHS), the variable that identifies ethnic groups in that dataset, the type of ethnic marker (language, ethnic group, mother tongue), as well as year or survey-round identifiers where appropriate.

```{r}
# Retrieve dataset dictionary
list.dict <- leda$get_list_dict()

# Show first entries
head(list.dict)

# All data types
unique(list.dict$type)

```


## Link data sets

Once familiar with the lists of ethnic groups that are part of the LEDA object, we can proceed to link the groups contained in any two lists of groups to each other. The LEDA object includes three methods to link lists of ethnic groups to each other, each of them described below.

### Link via set relations

We can first link lists *A* to lists *B* by analyzing the set of nodes on the language tree that groups *a* and *b* share. In the example below, we link two groups to each other as soon as they are associated with at least one common dialect on the language tree (`link.level = "dialect"`). As one specifies link levels closer to the root of the language tree, i.e. by setting `link.level = "language"` or `link.level = 5` (language tree level 5 of 16), the number of groups *b* linked to *a* increases and links become less precise.

The lists entered for parameters `lists.a` and `lists.b` offer a flexible way to select the lists of ethnic groups that are linked to each other. Note that you can enter any parameter combination that identifies at least one list of ethnic groups, but potentially many. The latter is helpful if you want to, for example, link all Afrobarometer surveys to the Ethnic Power Relations (EPR) data. It is generally (but not always) sensible to only link lists of ethnic groups within the same country borders by setting `by.country = T`.


```{r}

## Link all Afrobarometer groups (rounds 1-5) in Uganda to the FRT data.
setlink <- leda$link_set(lists.a = list(type = c("Afrobarometer"),  
                                        iso3c = c("UGA"),
                                        round = 4, marker = "language"),
                                lists.b = list(type = c("FRT"),
                                               iso3c = c("UGA")),
                                link.level = "dialect",
                                by.country = T,
                                drop.a.threshold = 0,
                                drop.b.threshold = 0,
                                drop.ethno.id = T)

## Have a look
head(setlink[, c("a.group", "b.group", "a.type", "b.type")])
```

One can further refine the link by constraining the arguments `drop.a.threshold` and `drop.b.threshold` that control the shares of common languages associated with groups *a* and *b* for a link to be realized. For eaxample, setting `drop.a.threshold = .5` ensures that in each link the language nodes of group *b* cover more than 50 percent of the language nodes associated with *a*. Conversely, setting `drop.b.threshold = .5` will ensure that in each pair of linked group *a* and *b*, group *a* covers more than 50 percent of the language nodes of *b*. More complex set relations can be implemented by setting the thresholds to 0 and switching `drop.ethno.id = FALSE`. The returned link table will then have multiple rows per linked pair of groups *a* and *b*, each coming with the ID of the language node they share.


### Link via linguistic distances

We can also make direct use of the language tree and link groups in lists *A* and *B* on the basis of their linguistic distances to each other. To do so, LEDA calculates linguistic distances first and then subsets the distance matrix to return the links queried by the user.

#### Compute linguistic distance between groups

The algorithm computes the full linguistic distance matrix between groups in lists *A* and *B*. Via the parameter `level`, users can specify whether they want links to be based on distances between ethnic groups' `"language"` or `"dialect`. As before, it is sensible to not link lists across country borders by setting `by.country = T`.

The linguistic distance between two languages or dialects $L_1$ and $L_2$ is computed as :

$1 - ((d(L_1,R) + d(L_2,R) - d(L_1,L_2))/(d(L_1,R) + d(L_2,R))))^{\delta}$

where $d(L_i,R)$ is the length of path from a language to the tree's origin and  $d(L_1,L_2)$ is the length of the shortest path from the first to the second language. $\delta$ is an exponent to discount short distances on the tree, reflected in the parameter `delta` below. Lastly, there are two ways to locate languages and dialects on the language tree. In the first, languages that are immediate children of a node that is located at level 4 of the language tree remain at their original level 5 (`expand = FALSE`).  In the second way, the tree is expanded, and all languages are located on level 15 and all dialects on level 16. This expansion of the tree naturally changes computed linguistc distances.

Because ethnic groups are often linked to multiple languages or dialects, there can be multiple linguistic distances between any group *a* and *b*. `agg_fun.a` and `agg_fun.b` control the aggregation of these distances. `agg_fun.a` determines for any language node in *a* how its distances to nodes of *b* are aggregated. `agg_fun.b` controls how the resulting distances between nodes in *a* and group *b* are aggregated to arrive at a single distance between *a* and *b*.

```{r}
## Compute distances
distance.df <- leda$ling_distance(lists.a = list(type = c("Afrobarometer"),
                                                 iso3c = "UGA",
                                                 round = 4, marker = "language"),
                                  lists.b = list(type = c("FRT"), iso3c = "UGA"),
                                  level = "dialect", by.country = T,
                                  delta = .5, expand = FALSE,
                                  agg_fun.a = min, agg_fun.b = min)

## Have a look
head(distance.df[, c("a.group", "b.group", "a.type", "b.type", "distance")])
```

### Link to closest linguistic neighbours

Based on the linguistic distances computed as discussed above, users can query, for every group *a* in lists *A* and for every list *B*, the closest linguistic neighbor *b*. Note that more than one nearest linguistic neighbor is returned wherever two or more closest groups *b* have the exact same lingusitic to *a*.


```{r}
mindistlink <- leda$link_minlingdist(lists.a = list(type = c("Afrobarometer"),  
                                                    iso3c = "UGA",
                                                    round = 4, marker = "language"),
                                     lists.b = list(type = c("FRT"), iso3c = "UGA"),
                                     level = "dialect",
                                     by.country = T,
                                     expand = FALSE,
                                     delta = .5,
                                     agg_fun.a = min, agg_fun.b = min)

## Have a look
head(mindistlink[, c("a.group", "b.group", "a.type", "b.type", "distance")])

```





### Link within linguistic distance


Instead of focusing on nearest linguistic neighbors only, users can also query, for every group *a* in lists *A* and for every list *B*, those groups *b* that fall within a specified distance `max.distance` of group *a*.

```{r}
withindistlink <- leda$link_withinlingdist(lists.a = list(type = c("Afrobarometer"),  
                                                          iso3c = "UGA",
                                                          round = 4, marker = "language"),
                                           lists.b = list(type = c("FRT"), iso3c = "UGA"),
                                           level = "dialect", max.distance = .1,
                                           by.country = T,
                                           delta = .5, expand = FALSE,
                                           agg_fun.a = min, agg_fun.b = min)

## Have a look
head(withindistlink[, c("a.group", "b.group", "a.type", "b.type", "distance")])
```





## Inspect coding of the ethnic group <--> language link

Sometimes, one might want to inspect the origins of a link between to groups. LEDA allows that by giving access to the entire raw data that underlies each match. You can query the link between any list of groups and the language tree with the following method.

The resulting table contains one column `link` that contains the language tree nodes linked to any group. Note that in cases of multiple links, they are separated by a '|'. In most cases, the level of a node on the language tree is indicated in squared brackets behind the nodes name. L1 to L14 indicate super-languages, 'lang' denotes languages, 'iso' language isocodes, and 'dial' refers to dialects.


```{r}
## Query raw link data
raw_ethno_links <- leda$get_raw_ethnolinks(param_list = list(type = "Afrobarometer",
                                                             round = 4,
                                                             marker = "language",
                                                             iso3c = "UGA"))

## Have a look
head(raw_ethno_links[, c("type","group", "link")])

```




## Add new links from groups to language tree

Having gained familiarity with the available ethnic links and methods, users can go a step further and link new lists of ethnic groups to the language tree. Doing so allows to link the new list of ethnic groups to every other list of ethnic groups covered by LEDA or independently added before.

### Prepare new links between ethnic groups and the tree

 First, one has to hand-code the link between ethnic groups and the language tree. However, this may be less tedious than it sounds. Via the method `LEDA$prepare_newlink_table()` one can access automatically generated suggestions to which language node(s) a particular group may link. These suggestions are generated via a fuzzy string match of a group's name to the names of (1) language nodes themselves, and (2) the names of ethnic groups already matched to the language tree. Thus, with every additional list of ethnic groups added to the data, linking new ones to the language tree becomes easier.


Once generated as shown below, the link table should be saved and the final links between ethnic groups and language nodes established by hand. I.e., users have to fill in the column `link`, using the information from the automatically generated suggestions, as well as secondary sources.

```{r}

## Make or load some dataset of ethnic groups
new.groups.df <- data.frame(group_name = c("Alur", "Iteso", "Kakwa"),
                            iso3c = c("UGA"),
                            marker = "ethnic group",
                            stringsAsFactors = F)


## Prepare a new link table
##   This table contains suggested links between each ethnic group
##   and language nodes. The columns "link", "comment", and "source"
##   have to be filled by hand and correspond to the final link to
##   a set of language nodes (separated by '|'), comments on the link,
##   and a source (if required).
newlink.df <- leda$prepare_newlink_table(group.df = new.groups.df,
                           groupvar = "group_name",
                           by.country = TRUE,
                           return = TRUE,
                           save.path =  NULL, overwrite = T,
                           prev_link_param_list = NULL,
                           levenshtein.threshold = .2,
                           levenshtein.costs = c(insertions = 1,deletions = 1, substitutions = 1))

newlink.df

```




### Add new links to a LEDA object  

Having hand-coded the link between the new list of ethnic groups and the language tree, one can now add the new list of groups to the LEDA object. The list now enters the object in the same manner as all 'native' LEDA lists, as well as any lists added beforehand.

```{r}

## First we need to encode links to the lanugage tree:
newlink.df$link[newlink.df$group == "Alur"] <- "Alur [L9]"
newlink.df$link[newlink.df$group == "Iteso"] <- "Teso [L7]"
newlink.df$link[newlink.df$group == "Kakwa"] <- "Kakwa [org]"
newlink.df$comment[newlink.df$group == "Kakwa"] <- "Kakwa same language as Bari, differs between language datasets."


## Add to LEDA
leda$add_tree_links(tree.link.df = newlink.df,
                    idvars = c("iso3c", "marker"),
                    type = "My Survey")

## Check type list
print(unique(leda$get_list_dict()$type))
```

For full traceability, the newly coded data is now also available in the raw data attached to LEDA and can be queried accordingly:

```{r}
## Query raw link data
raw_ethno_links <- leda$get_raw_ethnolinks(param_list = list(type = "My Survey"))

## Have a look
head(raw_ethno_links[, c("type","group", "link")])

```


### Join own data with other ethnic group lists

The new list can now be linked to any other list of ethnic groups in the LEDA object, in the same way as discussed above.

```{r}

## Get set link from my survey to FRT
setlink <- leda$link_set(lists.a = list(type = c("My Survey"), iso3c = "UGA"),
                         lists.b = list(type = c("FRT"), iso3c = "UGA"),
                         link.level = "dialect", by.country = T,
                         drop.a.threshold = 0, drop.b.threshold = 0)

## Have a look
head(setlink[, c("a.group", "b.group", "a.type", "b.type")])
```

### Submit new lists to LEDA project

Given that the value of LEDA increases exponentially with the number of lists available in the R-package, we would greatly appreciate if you could share any new lists that you link to the language tree. New lists can be new rounds of survey data (e.g. Afrobarometer, DHS) or any list of ethnic groups that is based on publicly available data. You can do so by sending us an email to carlvs /at/ ethz.ch or opening an issue with the attached link file via LEDA's Github page. Shared link files should have the format returned by the method `LEDA$prepare_newlink_table()` and have the `link` column filled wherever possible. 
